---
output:
  html_document:
    toc: no
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
syl.comp <- read.csv("Syllabus Components.csv")
require(dplyr)
```

# `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Title")])`

Date: `r format(Sys.time(), '%d %B %Y')`

Instructor: `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Instructor")])`

Office: `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Office")])`

Office Hours: `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Office Hours")])`

Email: `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Email")])`

Website: https://rtleaf.github.io/rtleaf-COA-606-Statistics-for-Scientists-and-Engineers/

Course Meeting Day and Time: `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Meeting Days")])`, `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Meeting Time")])`

## Course Description and Objectives
This course addresses basic approaches to experimental design, statistical analysis, and presentation of quantitative information.

## At the conclusion of this course: 

By the end of this course, you will be able to:

*	Define what comprises the field of statistical analysis.
*	What is meant by random samples, random sampling, and understanding how these aspects of sampling are critical for description and inference.
*	Explain the differences and similaries among variables, statistics, and parameters.
*	Define a cumulative density function and a probability density function and describe several common distributional families (e.g., normal, binomial, chi-square).
*	Understand and be familiar with the applications of:
*	Frequency distributions
*	Power analysis
*	Summary statistics as a tool for describing data
*	Means comparison techniques for testing hypotheses
*	ANOVA comparison techniques for testing hypothesis
*	Linear regression techniques for describing relationships between random variables.
*	Make appropriate decisions as a part of a statistical data analysis.

## Course Materials
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Materials 1")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Materials 2")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Materials 3")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Materials 4")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Materials 5")])`

## Course Scheduling
```{r, echo=FALSE, message=FALSE}
gs.scale <- read.csv("course_scheduling.csv")
gs.scale <- gs.scale[,1:5]
names(gs.scale) <- c("Day", "Section", "Activity", "Reading", "Link")
knitr::kable(gs.scale)
```

## Course Workload Statement
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Course Workload Statement")])`

## Course Evaluation
```{r, echo=FALSE, message=FALSE}
gs.scale <- read.csv("Grading Scale.csv")
names(gs.scale) <- c("Percentage", "Letter Grade")
knitr::kable(gs.scale)
```

## Course Evaluation - Participation and Attendence
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Participation and Attendence")])`

## Homework policy
*	There is no due date for the HW, you may turn it in anytime. 

*	Graded HW will be returned withing 5 working days of its receipt. 

* You will upload your scanned .pdf HW labled with your last name and assignment (example "Leaf_HW01.pdf", "Smith_HW02.pdf", "Jones_HW03.pdf"). Each HW assignment will be comprised of a single and labeled PDF file.

* Load your files to this folder:
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "HW.Link")])`

*	However, no work will be accepted after 5 PM CST 12/10/2021. The upload window will close at that time.

## Grading scale
```{r, echo=FALSE, message=FALSE}
course.eval <- read.csv("Course Evaluation.csv")
names(course.eval) <- c("Evaluation type",	"Number",	"Points per item",	"Total points")
knitr::kable(course.eval)
```

## Exam Bank
The bank of exams and study material can be found here:
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "ExamBank")])`

## COVID-19

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 01")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 02")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 03")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 04")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 05")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 06")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 07")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 08")])`

* `r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Health Protocol 09")])`

## Academic Support Resources
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Academic Support Resources")])`

## Mental Well-Being Statement
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Mental Well-Being Statement")])`

## Nondiscrimination Statement
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Nondiscrimination Statement")])`

## Confidentiality and Mandatory Reporting
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Confidentiality and Mandatory Reporting")])`

## Academic Integrity
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Academic Integrity")])`